Permutation Pattern matching in\\ $(213,231)$-avoiding permutations

### 2016-10-12 article.tex
revision

\begin{abstract}
Given permutations $\sigma$ of size $k$ and $\pi$ of size $n$ with $k<n$, the
\emph{permutation pattern matching} problem is to decide whether $\sigma$ occurs
$\pi$ as an order-isomorphic subsequence.
We give a linear-time algorithm in case both $\pi$ and $\sigma$ avoid
the two size-$3$ permutations $213$ and $231$.
For the special case where only $\sigma$ avoids $213$ and $231$, we present a
$O(max(kn^2,n^2\log(\log(n)))$-time algorithm. 
We extend our research to bivincular patterns that avoid $213$ and $231$ and present a $O(kn^4)$-time algorithm.
Finally we look at the related problem of the longest subsequence which avoids $213$ and $231$.  
\end{abstract}

@@

A permutation $\sigma$ is said to occur in another permutation $\pi$,
denoted $\sigma \preceq \pi$,
if there exists a subsequence of elements of $\pi$ that has the same relative
order as $\sigma$.
Otherwise, $\pi$ is said to \emph{avoid} the permutation $\sigma$.
For example the permutation $123$ (resp. $321$) occurs in a second permutation if
the second permutation has an increasing (resp. a decreasing) subsequence of size $3$.
Similarly, $213$ occurs in $6152347$, but $231$ does not occurs in $6152347$.
During the last decade, the study of patterns in permutations has
become a very active area of research \cite{Kitaev:book:2011} and
a whole annual conference (\textsc{Permutation Patterns}) focuses
on patterns in permutations.

We consider here the so-called \emph{permutation pattern matching} problem
(also sometimes referred to as the \emph{pattern involvement problem}):
Given two permutations $\sigma$ of size $k$ and $\pi$ of size $n$, the problem is to decide whether
$\sigma \preceq \pi$ (the problem is attributed to Wilf in \cite{Bose:Buss:Lubiw:1998}).
The permutation pattern matching is known to be \NPhard~\cite{Bose:Buss:Lubiw:1998}.
It is however polynomial-time solvable by brute-force enumeration
if $\sigma$ has bounded size.
Improvements to this algorithm were presented in
\cite{Albert:Aldred:Atkinson:Holton:ISAAC:2001} and
\cite{Ahal:Rabinovich:2008},
the latter describing a $O(n^{0.47k+o(k)})$-time algorithm.
Bruner and Lackner \cite{DBLP:journals/corr/abs-1204-5224}
gave a fixed-parameter algorithm solving the permutation pattern matching problem with
an exponential worst-case runtime of $O(1.52^{n}.n.k)$.
This is an improvement upon the $O(k {n \choose k})$ runtime required by
brute-force search without imposing restrictions on $\sigma$ and $\pi$,
where one has to enumerate all the ${n \choose k}$ different subsequences of size $k$ in $\pi$
and test if one of them has the same relative
order as $\sigma$.  
A recent major step was taken by Marx and Guillemot
\cite{Guillemot:Marx:SODA:2014}.
They showed that
the permutation pattern matching is fixed-parameter tractable (FPT) for by the size of $\sigma$.

A few particular cases of the permutation pattern matching problem have been attacked successfully.
The case of any increasing pattern is solvable in 
$O(n \log \log k)$ time in the RAM model \cite{Crochemore:Porat:2010},
improving the previous 30-year bound of $O(n \log k)$.
% (The algorithm also improves on the previous
% $O(|\pi| \log \log |\pi|)$ bound.)
Furthermore, the patterns $132$, $213$, $231$, $312$ can all be handled in linear-time
by stack sorting algorithms.
Any pattern of size $4$ can be detected in $O(n \log n)$ time
\cite{Albert:Aldred:Atkinson:Holton:ISAAC:2001}.
Algorithmic issues for $321$-avoiding patterns matching for permutations have been investigated in
\cite{Guillemot:Vialette:ISAAC:2009} and more recently in \cite{2015arXiv151006051A}.
The permutation pattern matching problem is also solvable in
polynomial-time for separable patterns \cite{Ibarra:1997,Bose:Buss:Lubiw:1998}
(see also \cite{Bouvel:Rossin:Vialette:CPM:2007} for the longest common subsequence problem and related ones on separable permutations).
Separable permutations are permutations that do not have an occurrence of
$2413$ nor $3142$, and they are enumerated by the Schr{\"o}der numbers.
%	(sequence A006318 in OEIS).
%	To the best of our knowledge,
%	separable permutations first arose in the work of
%	Avis and Newborn~\cite{Avis:Newborn:1981},
%	who showed that they are precisely the permutations which can be sorted by an
%	arbitrary number of pop-stacks in series,
%	where a pop-stack is a restricted form of stack in which any pop operation
%	pops all items at once.
(Notice that the separable permutations include as a special case the
stack-sortable permutations, which avoid the pattern $231$.)

>>>>

\begin{abstract}
Given permutations $\sigma$ of size $k$ and $\pi$ of size $n$ with $k<n$, the
\emph{permutation pattern matching} problem is to decide whether $\sigma$ occurs
$\pi$ as an order-isomorphic subsequence.
We give a linear-time algorithm in case both $\pi$ and $\sigma$ avoid
the two size-$3$ permutations $213$ and $231$.
For the special case where only $\sigma$ avoids $213$ and $231$, we present a
$O(max(kn^2,n^2\log \log n)$-time algorithm.
We extend our research to bivincular patterns that avoid $213$ and $231$ and present a $O(kn^4)$-time algorithm.
Finally we look at the related problem of the longest subsequence which avoids $213$ and $231$.
\end{abstract}

@@

A permutation $\sigma$ is said to \emph{occur} in another permutation $\pi$
(or $\pi$ \emph{contains} $\sigma$),
denoted $\sigma \preceq \pi$,
if there exists a subsequence of elements of $\pi$ that has the same relative
order as $\sigma$.
Otherwise, $\pi$ is said to \emph{avoid} the permutation $\sigma$.
For example a permutation contains the permutation $123$ (resp. $321$) if
it has an increasing (resp. a decreasing) subsequence of size $3$.
Similarly, $213$ occurs in $6152347$, but $231$ does not occurs in $6152347$.
During the last decade, the study of patterns in permutations has
become a very active area of research \cite{Kitaev:book:2011} and
a whole annual conference (\textsc{Permutation Patterns}) focuses
on patterns in permutations.

We consider here the so-called \emph{permutation pattern matching} problem
(also sometimes referred to as the \emph{pattern involvement problem}):
Given two permutations $\sigma$ of size $k$ and $\pi$ of size $n$, the problem is to decide whether
$\sigma \preceq \pi$ (the problem is attributed to Wilf in \cite{Bose:Buss:Lubiw:1998}).
The permutation pattern matching is known to be \NPhard~\cite{Bose:Buss:Lubiw:1998}.
It is however polynomial-time solvable by brute-force enumeration
if $\sigma$ has bounded size.
Improvements to this algorithm were presented in
\cite{Albert:Aldred:Atkinson:Holton:ISAAC:2001} and
\cite{Ahal:Rabinovich:2008},
the latter describing a $O(n^{0.47k+o(k)})$-time algorithm.
Bruner and Lackner \cite{DBLP:journals/corr/abs-1204-5224}
gave a fixed-parameter algorithm solving the permutation pattern matching problem with
an exponential worst-case runtime of $O(1.52^{n}.n.k)$.
This is an improvement upon the $O(k {n \choose k})$ runtime required by
brute-force search without imposing restrictions on $\sigma$ and $\pi$,
where one has to enumerate all the ${n \choose k}$ different subsequences of size $k$ in $\pi$
and test if one of them has the same relative
order as $\sigma$.
Guillemot and Marx \cite{Guillemot:Marx:SODA:2014} showed that the permutation pattern matching problem
is solvable in $2^{O(k^2 \log k)}$, and hence is fixed-parameter tractable with respect to the length $k$
of the pattern $\sigma$ (standard parameterization).
However, Mach proved that the permutation containment problem under the
standard parameterization does not have a polynomial size kernel (assuming $\NPclass \not\subseteq \coNPclass / \POLY$
\cite{Mach:thesis:1992}.

A few particular cases of the permutation pattern matching problem have been attacked successfully.
The case of  increasing patterns is solvable in
$O(n \log \log k)$-time \cite{Crochemore:Porat:2010},
improving the previous 30-year bound of $O(n \log k)$.
% (The algorithm also improves on the previous
% $O(|\pi| \log \log |\pi|)$ bound.)
Furthermore, the patterns $132$, $213$, $231$, $312$ can all be handled in linear-time
by stack sorting algorithms.
Any pattern of size $4$ can be detected in $O(n \log n)$ time
\cite{Albert:Aldred:Atkinson:Holton:ISAAC:2001}.
Algorithmic issues for $321$-avoiding patterns matching for permutations have been investigated in
\cite{Guillemot:Vialette:ISAAC:2009} and more recently in \cite{2015arXiv151006051A}.
The permutation pattern matching problem is also solvable in
polynomial-time for separable patterns \cite{Ibarra:1997,Bose:Buss:Lubiw:1998}
(see also \cite{Bouvel:Rossin:Vialette:CPM:2007} for the longest common subsequence problem and
related ones on separable permutations).
Separable permutations are permutations that do not have an occurrence of
$2413$ nor $3142$, and they are enumerated by the Schr{\"o}der numbers.
%	(sequence A006318 in OEIS).
%	To the best of our knowledge,
%	separable permutations first arose in the work of
%	Avis and Newborn~\cite{Avis:Newborn:1981},
%	who showed that they are precisely the permutations which can be sorted by an
%	arbitrary number of pop-stacks in series,
%	where a pop-stack is a restricted form of stack in which any pop operation
%	pops all items at once.
(Notice that the separable permutations include as a special case the
stack-sortable permutations, which avoid the pattern $231$.)
The major negative result is given in \cite{DBLP:journals/corr/JelinekK16}.
They prove that
the permutation pattern matching problem is \NPC
when $\sigma$ has no decreasing subsequence of length $3$ and
$\pi$ has no decreasing subsequence of length $4$.
This provides the first known example of the permutation pattern matching problem
being hard when one or both of $\pi$ and $\sigma$ are restricted to a proper hereditary
class of permutations.
